\chapter{Timed Systems and Semantics}\label{chap:2} 
\minitoc
\section{Timed Transition Systems}
Transition systems provide a general and convenient method for modeling 
systems and have been used frequently to model the behavior of 
software and hardware systems. They define graphs where nodes
represent the possible \emph{states} of the system, and edges model 
\emph{transitions}, that is, state changes. A state encodes all the relevant 
information at a certain instant, whereas a transition describes how the system
evolves between two states. 

Nowadays, several variants of transition system formalisms have been 
proposed. For instance, a labeled transition system is a transition system
where the set of transitions is labeled by \emph{actions}.
In this thesis, we use \emph{Timed Transition Systems} (TTS)
to explicitly model the effect of time passage (besides the actions)
on the states of the system. Formally, it is defined as follows:  

\begin{definition}[Timed Transition System]\label{def:tts}
  A timed transition system is a tuple $\TTS$ such that $\mc{Q}$ is 
  a set of states, $q_0\in\mc{Q}$ is the initial state, $\Sigma$ is a set of
  actions, $\mb{K}$ is a time domain and $\to\subseteq\mc{Q}\times
  (\Sigma\cup\mb{K})\times\mc{Q}$ is the transition relation.
\end{definition}
Consequently, we distinguish two types of transition:
\begin{itemize}
  \item \emph{action step} $\transit{\sigma}$ for $\sigma\in\Sigma$ and we write 
    $q\transit{\sigma}q'$. 
  \item \emph{time step} $\transit{d}$ for $d\in\mb{K}$ and we write $q\transit{d}q'$. 
\end{itemize}
We write $q\transit{d,\sigma}q''$ if there exists $q'\in\mc{Q}$
such that $q\transit{d}q'\transit{\sigma}q''$. We say that $q'$ is
a time successor of $q$.

Given a TTS $\TTS$, a \emph{run} $\exec$ of $\tts$ (also called 
\emph{execution sequence}), is a path that alternates action steps and 
time steps, that is:
\begin{displaymath}
  \exec=s_0\sigma_0 s_1\sigma_1 s_2\cdots \text{ such that } 
  s_i\in\mc{Q}, \  s_i\transit{\sigma_{i}}s_{i+1},\text{ and } 
  i\in\integerpoz,\sigma_i\in\Sigma\cup\mb{K}
\end{displaymath}

We denote by $\texeci$ the total elapsed time until point $i$, that is,
$\sum_{j<i}\sigma_j$ such as $\sigma_j\in\mb{K}$. In the same way, $\texec$ represents
the total elapsed time during $\varrho$, and is defined to be the limit of $\texeci$
if the sequence converges and $\infty$ otherwise.
A run $\exec$ is said to be an \emph{initial run} if $s_0=q_0$. 

A state $q\in\mc{Q}$ is called reachable if there exists an initial run
that leads to state $q$. We put $Reach(\tts)$ to denote the set of all 
reachable states of $\tts$.
\subsection{Comparing Timed Transition Systems}
In this thesis, we use the concept of (bi)simulation~\cite{bsim} 
in order to attest the similarity of timed transition systems.
\begin{definition}[Simulation]\label{def:sim}
  Given two TTS $\TTSb{1}$ and $\TTSb{2}$, a simulation relation from 
  $\ttsb{1}$ to $\ttsb{2}$ is a binary relation $\mc{R}\subseteq\mc{Q}_1
  \times\mc{Q}_2$ such that:
  \begin{itemize}
    \item $\forall(q_1,q_2)\in\mc{R},\forall\sigma\in\Sigma,\ 
      q_1\transitb{\sigma}{1}q_1'\Rightarrow\exists q'_2\in\mc{Q}_2 
      \text{ such that } q_2\transitb{\sigma}{2}q_2'\wedge
      (q_1',q_2')\in\mc{R}$
    \item $\forall(q_1,q_2)\in\mc{R},\forall d\in\mb{K},\ 
      q_1\transitb{d}{1}q_1'\Rightarrow\exists q'_2\in\mc{Q}_2 
      \text{ such that } q_2\transitb{d}{2}q_2'\wedge
      (q_1',q_2')\in\mc{R}$
  \end{itemize}
\end{definition}
  $\ttsb{2}$ simulates $\ttsb{1}$, denoted by $\ttsb{1}\simu{\mc{R}}\ttsb{2}$
  means that $\ttsb{2}$ can do everything $\ttsb{1}$ does. Notice that if
  $\ttsb{1}\simu{\mc{R}}\ttsb{2}$ and $\ttsb{2}\simu{\mc{R}}\ttsb{1}$,
  we say that $\ttsb{1}$ and $\ttsb{2}$ are bisimilar, denoted by
  $\ttsb{1}\eqv{\mc{R}}\ttsb{2}$.  

In some cases, this notion of simulation is refined in order to consider only
a part of a system behavior. This is usually the case when a system
performs \emph{internal} (or \emph{silent}) actions not visible by external
observers. This variant of simulation is called \emph{weak simulation}.

\begin{definition}[Weak Simulation]\label{def:wsim}
  Given two TTS $\TTSbw{1}$ and $\TTSbw{2}$, where $\tau$ actions 
  represent silent (unobservable) actions, a weak simulation relation from 
  $\ttsb{1}$ to $\ttsb{2}$, denoted $\ttsb{1}\simuw{\mc{R}}\ttsb{2}$, 
  is a binary relation $\mc{R}\subseteq\mc{Q}_1
  \times\mc{Q}_2$ such that:
  \begin{itemize}
    \item $\forall(q_1,q_2)\in\mc{R},\ 
      q_1\transitb{\tau}{1}q_1'\Rightarrow\exists q'_2\in\mc{Q}_2 
      \text{ such that } q_2\transitb{\tau^*}{2}q_2'\wedge
      (q_1',q_2')\in\mc{R}$
    \item $\forall(q_1,q_2)\in\mc{R},\forall\sigma\in\Sigma,\ 
      q_1\transitb{\sigma}{1}q_1'\Rightarrow\exists q'_2\in\mc{Q}_2 
      \text{ such that } q_2\transitb{\tau^*\sigma\tau^*}{2}q_2'\wedge
      (q_1',q_2')\in\mc{R}$
    \item $\forall(q_1,q_2)\in\mc{R},\forall d\in\mb{K},\ 
      q_1\transitb{d}{1}q_1'\Rightarrow\exists q'_2\in\mc{Q}_2 
      \text{ such that } q_2\transitb{\tau^*d\tau^*}{2}q_2'\wedge
      (q_1',q_2')\in\mc{R}$
  \end{itemize}
\end{definition}
  We say that $\ttsb{1}$ and $\ttsb{2}$ are observationally equivalent,
  denoted $\ttsb{1}\eqvw{\mc{R}}\ttsb{2}$ , if it exists a weak simulation 
  from $\ttsb{1}$ to $\ttsb{2}$ and vice versa.
  
  In chapter~\ref{chap:6}, we will introduce an even weaker notion of simulation 
  that characterizes the degree of closeness (in term of delays)
  between two timed systems. 

\subsection{Reactive Timed Systems}
Reactive systems are supposed to execute forever, that is,
they are supposed to act infinitely often. We refer to this characteristic as 
the \emph{requirement of progress}~\cite{progress}. 
Particularly, a timed system evolves either through an action step or 
by letting the time pass (time step).
This evolution imposes thus two requirements of progress:
discrete progress (resp. time progress) meaning that a timed system
should be able to perform action steps (resp. time steps) infinitely
often. In the physical world however, no matter how fast a system can
evolve, it cannot be infinitely fast. This induces the following constraints
on the progress of time:
\begin{enumerate}
  \item Only a finite number of actions can happen in a certain amount of time
  \item Only a bounded number of actions can happen in zero time
\end{enumerate}

\subsubsection{Time Progress (Zeno runs \& Timelocks)} 
We distinguish two types of anomalies that infringe the time progress in
a timed system namely \emph{zeno} runs and \emph{timelocks}.
A run $\exec$ is called \emph{zeno} if it is 
an infinite run and if $\texec\neq\infty$. Such a run transgresses the 
first point of the time progress presented above.
Timelocks are states from which all infinite runs starting from these states
are zeno.
\subsubsection{Discrete Progress (Deadlocks)}
States violating the discrete progress are called \emph{deadlock} states.
Formally, a state is said to be deadlock if no action can be executed from 
that state nor from any of its time successors. 

Any model of a reactive timed system should properly capture the behavior 
of the latter. Particularly, the corresponding model must react infinitely often,
that is, it must not block time or execute an unbounded number of actions in zero 
time. In that sense, we can say that deadlocks and timelocks are modeling errors
that needs to be cleared, either during the modeling process (which is tedious 
for large scale systems) or by providing verification methods that guarantee
their absence. 


\section{Timed Systems Syntax and Semantics}\label{sec:2.2}
\subsection*{Clocks}
In order to represent and measure the dense time domain, we rely on positive 
real valued variables, the \emph{clocks}. Clocks are 
positive real variables increasing synchronously (with the same rate) in a 
given system. They are used to express the progress of time and impose 
a certain dynamic on the execution of a timed system.

Given a finite set of clocks $\mc{X}$, we define the valuation function
$\val:\mc{X}\to\realpoz$ assigning to each clock $x$ a positive real value 
$\val(x)$. We put $\Val$ to denote the set of all valuations.
For a valuation $\val\in\Val$ and $d\in\realpoz$, $\val+d$ is the valuation
satisfying $(\val+d)(x)=\val(x)+d$, while for a subset of clocks 
$r\subseteq\mc{X}$, $\val[r]$ is the valuation obtained from $\val$ by 
resetting clocks of $r$, that is, $\val[r](x)=0$ for $x\in r$ and
$\val[r](x)=\val(x)$ otherwise. We write {\bf 0} for the valuation 
that assigns 0 to every clock.

An \emph{atomic clock constraint} is an expression of the form:
\begin{displaymath}
  c:=\true \ | \ x\#k \ |\ x-y\#k \
\end{displaymath}
where $x$ and $y$ are clocks in $\mc{X}$, $\#\in\{<,\le,=,\ge,>\}$,
and $k\in\integer$. A clock constraint is a conjunction of atomic clock 
constraint, that is:
\begin{equation}\label{eq:cc}
  c:=\true \ | \ x\#k \ |\ x-y\#k \ | \ c_1\wedge c_2 
\end{equation}
with $c_1$ and $c_2$ being atomic clock constraints. We write $\mc{C(X)}$
for the set of clock constraints over $\mc{X}$.
Given a clock constraint $c$ and a valuation $\val$, we say that $\val$
satisfies $c$, denoted $\val\models c$, if all constraints are satisfied
when each $x\in\mc{X}$ is replaced by $\val(x)$. 
We also consider for a clock constraint $c$ the classical \emph{backward} 
and \emph{forward} operators on
clock constraints:
\begin{align*}
  &\text{ \bf Backward } &:& \val\models\backward c\Leftrightarrow 
  \exists d\in\realpoz.\ \val+d\models c\\ 
  &\text{ \bf Forward } &:& \val\models\forward c\Leftrightarrow 
  \exists d\in\realpoz.\ \val-d\models c\\ 
\end{align*}

Additionally, we also use another variant of the backward operator considering
lower bounds $l\in\integerpoz$ and upper bounds $u\in\integerpoz\cup
\{+\infty\}$:

\begin{displaymath}
  \val\models\backwardp{l}{u} c\Leftrightarrow 
  \exists d\in\realpoz, l\le d\le u. \ \val+d\models c\\ 
\end{displaymath}

\subsection{Timed Components Syntax and Semantics}
In this thesis, components are timed automata and systems are compositions of timed automata
with respect to multiparty interactions. The timed automata we use are essentially the ones from
~\cite{AlurD94}, however, slightly adapted to embrace a uniform notation the dissertation.
\begin{definition}[Timed Component]\label{def:tc}
  A component is a tuple $B=(\Loc,\loc_0,\X,\A,\E,\I)$, where:
  \begin{itemize}
    \item $\Loc$ is a finite set of locations, with $\loc_0\in\Loc$ is the 
      initial location,
    \item $\mc{X}$ is a finite set of clocks,
    \item $\mc{A}$ is a finite set of actions
    \item $\mc{E}\subseteq\Loc\times(\mc{A}\times\mc{C(X)}\times2^{\mc{X}})
      \times\Loc$ is a finite set of transitions labeled with an action, 
      a clock constraint (guard), and a set of clocks to be reset,
    \item $\mc{I}:\Loc\to\mc{C(X)}$ is the function assigning an invariant
      to each location. Notice that invariants are restricted to conjunction
      of atomic clock constraints of the form $x\le k$. 
  \end{itemize}
\end{definition}

Throughout this thesis, we consider 
\emph{deterministic} timed components, that is, at a given location
$\loc$ and for a given action $a$, there is up to one outgoing transition
from $\loc$ labeled by a. 
A transition $e=(\loc,(a,g,r),\loc')\in\mc{E}$ is also denoted by
$\loc\transit{a,g,r}\loc'$. We write $\source(e)$, $\target(e)$, $\action(e)$, $\guard(e)$
and $\reset(e)$ for $\loc$, $\loc'$, $a$, $g$ and $r$, respectively.
We also denote by $\Phi(a,\loc)$ the guard 
of the transition labeled by $a$ and outgoing from $\loc$
if it exists, and $\false$ otherwise. It is formalized as follows:
\[\Phi(a,\loc)=\begin{cases}
  g,& \text{ if } \exists e=(\loc,a,g,r,\loc')\in\mc{E}  \\
  \false,& otherwise
\end{cases}\]

\begin{example}
  Figure~\ref{fig:tc} depicts a timed component $B$ where locations
  are represented by circles and transitions are the directed arrows
  from a location to another. The initial location ($\loc_0$ here) is
  represented by a double circle. The component 
  $B=(\Loc,\loc_0,\mc{X},\mc{A},\mc{E},\mc{I})$ is defined such that:
  \begin{itemize}
    \item $\Loc=\{\loc_0,\loc_1\}$,
    \item $\mc{X}=\{x\}$,
    \item $\mc{A}=\{a,b\}$,
    \item $\mc{E}=\{e_1,e_2\}$ where
      \begin{itemize}
        \item $e_1=(\loc_0,a,2\le x\le4,\emptyset,\loc_1)$
        \item $e_2=(\loc_1,b,\true,\{x\},\loc_0)$
      \end{itemize}
    \item $\mc{I}$ assigns the following invariants for locations: 
      $\Inv{\loc_0}=x\le4$ and $\Inv{\loc_1}=\true$
  \end{itemize}
  Notice that when a clock constraint (respectively an invariant is not
  shown on a transition (respectively location) it is interpreted as $\true$.
  This is the case for transition $e_2$ and location $\loc_2$.
\end{example}
\input{Figures/tc.tex}

\begin{definition}[Standard Semantics]\label{def:std_sem}
The semantics of a timed component $B=(\Loc,\loc_0,\X,\A,\E,\I)$ is given by the timed transition
  system $\TTSs$ where:
  \begin{itemize}
    \item $\mc{Q}=\Loc\times\Val$ denotes the states
      of $B$ with $q_0=(\loc_0,0)$ being the initial state, 
    \item $\to\subseteq\mc{Q}\times(\mc{A}\cup\realpos)\times\mc{Q}$
    denotes the set of transitions between states according to the rules:
  \begin{itemize}
    \item $(\loc,\val)\transit{a}(\loc',\val[r])$ if $\loc\transit{a,g,r}
      \loc'$, $\val\models g$, and $\val[r]\models\Inv{\loc'}$
      (action step) 
    \item $(\loc,\val)\transit{d}(\loc,\val+d)$ if $\forall d'\in[0,d]$,
      $\val+d'\models \Inv{\loc}$ (time step) 
  \end{itemize}
  \end{itemize}
\end{definition}
Notice that since the invariants are restricted to conjunction of upper bound
atomic constraints, the time step can be simplified to:
\begin{displaymath}
  (\loc,\val)\transit{d}(\loc,\val+d)\text{ if }\val+d\models \Inv{\loc} 
\end{displaymath}

In this thesis, we always assume components with $\emph{well formed}$
guards, that is, for a transition $\loc\transit{a,g,r}\loc'$,
$\big(\val\models g\big)\Rightarrow\big(\val\models\Inv{\loc}\wedge\val[r]\models
\Inv{\loc'}\big)$ for any $\val\in\Val$. This ensures that the reachable states
always satisfy the location invariants. The rule on action step becomes then:
\begin{displaymath}
  (\loc,\val)\transit{a}(\loc',\val[r])\text{ if } \loc\transit{a,g,r}
      \loc'\text{ and } \val\models g 
\end{displaymath}
\begin{remark}
  When used in predicate definition, clock constraints are straightforwardly 
  applied to clock valuations of states and thus interpreted as $\true$
  or $\false$.
\end{remark}
We define the predicate $\enabled{a}$ characterizing states $(\loc,\val)$
from which an action $a$ is enabled, that is, such that $(\loc,\val)
\transit{a}(\loc',\val')$. It is formalized as follows:
\begin{displaymath}
  \enabled{a}=\bigvee_{\loc\in\Loc}\al{\loc}\wedge\Phi(a,\loc)
\end{displaymath}
where $\al{\loc}$ is $\true$ on states whose location is $\loc$.
In the same way, we define the predicates $\enabledbackward{a}$,
$\enabledbackwardb{a}{l}{u}$ and $\enabledforward{a}$ describing, respectively,
states from which an action $a$ can be executed after some time step,
some bounded time step or states that are time successors of states satisfying $\enabled{\alpha}$
(see Figure~\ref{fig:zones}). These predicates can be formally written 
as follows:
\begin{align*}
  &\enabledbackward{a}\hspace{1.8mm}=\bigvee_{\loc\in\Loc}\al{\loc}\wedge
  \backward\Phi(a,\loc)  \\
  &\enabledbackwardb{a}{l}{u}=\bigvee_{\loc\in\Loc}\al{\loc}\wedge
  \backwardp{l}{u}\Phi(a,\loc)\\
  &\enabledforward{a}\hspace{1.9mm}=\bigvee_{\loc\in\Loc}\al{\loc}\wedge
  \forward\Phi(a,\loc)
\end{align*}
\input{Figures/zone.tex}
A state $(\loc,\val)$ is said \emph{urgent} if time cannot progress from
this state, that is, $\nexists d\in\realpos$ such that $(\loc,\val)\transit{d}
(\loc,\val+d)$. Urgent states of a component B are characterized by the predicate:
\begin{equation}\label{eq:urg}
  \urgent(B)=\bigvee_{\loc\in\Loc}\al{\loc}\wedge urg(\loc)
\end{equation}
where $urg(\loc)$ is a clock constraint characterizing the valuations from
which time cannot progress with respect to the location invariant 
$\Inv{\loc}$, that is, assuming that 
$\Inv{\loc}=\bigwedge_{i=1}^m x_i\le k_i$ then 
$urg(\loc)=\bigwedge_{i=1}^m x_i\ge k_i$. Notice that due to well formed
guards, an urgent reachable state satisfies also Expression~\ref{eq:urg} if inequalities
$x_i\ge k_i$ on clocks are replaced by equalities $x_i=k_i$ in the expression
of $urg(\loc)$.

\begin{definition}[Strongly Non-Zeno Timed Component]\label{def:snz}
  Given a timed component $B$, a \emph{structural loop} of $B$ is a sequence 
  of distinct transition $e_1\cdots e_m$ such that $\forall i\in\{1,\cdots,m
  \},\target(e_i)= \source(e_{i+1})$. $B$ is called \emph{strongly non-zeno} 
  if for every structural loop there exists a clock $x$ and some $0\le i,j
  \le m$ such that:
  \begin{itemize}
    \item $x$ is reset in step $i$, that is, $x\in\reset(e_i)$
    \item $x$ is bounded from below in step j, that is, $(x<1)\wedge\guard(e_j)
      =\false$
  \end{itemize}
\end{definition}
Intuitively, this definition implies that at least 1 time unit elapses 
at each loop of $B$.

\begin{lemma}[\cite{progress}]\label{lm:snz}
  If $B$ is strongly non-zeno then every infinite run of $B$ is non-zeno
\end{lemma}

The following corollary is an immediate consequence of lemma~\ref{lm:snz}. 
\begin{corollary}\label{cr:snz}
Given a timed component $B$, if B is strongly non-zeno then it is 
  timelock free.
\end{corollary}

Corollary~\ref{cr:snz} highlights an interesting fact of strong non-zenoness.
It discharges from the trouble of checking time progress. In particular, checking
the progress of a timed system is reduced to checking its deadlock freedom.

\begin{definition}[Deadlocks]\label{def:dl}
  Given a timed component $B$. We say that a state $(\loc,\val)$ of $B$ is 
  a \emph{deadlock} if and only if no action can be executed from this state
  or any of its time successors, that is:
  \begin{displaymath}
    \neg\Big(\exists a\in\mc{A}. \ (\loc,\val)\transit{a}(\loc',\val')\vee
    \exists d>0. \ (\loc,\val)\transit{d}(\loc,\val+d)\transit{a}(\loc',\val')\Big)
  \end{displaymath}
  Deadlock states are characterized by the following predicate:
  \begin{displaymath}
   \bigvee_{\loc\in\Loc}\al{\loc}\wedge
 \neg\Big(\bigvee_{a\in\mc{A}}
    \backward\big(\enabled{a}\wedge\Inv{\loc}\big)\Big)
  \end{displaymath}
  Because of well formed guards, the above can be simplified into:
  \begin{equation}\label{eq:dl}
    \bigwedge_{a\in\mc{A}}\neg\enabledbackward{a}
  \end{equation}
\end{definition}
A deadlock $(\loc,\val)$ is called an \emph{action-time-lock} when no action can
execute nor time can progress from $(\loc,\val)$, that is:
  \begin{displaymath}
    \neg\Big(\exists a\in\mc{A}. \ (\loc,\val)\transit{a}(\loc',\val')\vee
    \exists d>0. \ (\loc,\val)\transit{d}(\loc,\val+d)\Big)
  \end{displaymath}
  Action-time-locks verifies the following predicate:
  \begin{equation}\label{eq:atl}
    \bigwedge_{a\in\mc{A}}\neg\enabled{a}\wedge\bigvee_{\loc\in\Loc}
    \big(\al{\loc}\wedge urg(\loc)\big)
  \end{equation}
\subsection{Timed Systems}
The common practice in component-based timed systems is to have several components
executing in parallel while their clocks increase synchronously. Moreover,
it is often mandatory to restrict the components behaviors so as to achieve 
a given global property. This is usually achieved by coordinating the execution of 
actions in components using synchronization mechanisms.
In what follows, components communicate by means
of \emph{multiparty interactions}. A multiparty interaction is a rendez-vous 
synchronization between actions of a fixed subset of components. It takes place 
only if all the participants agree to execute the corresponding actions. Given 
$n$ components $B_i$, with disjoint sets of actions $\mc{A}_i$, an interaction
is a subset of actions $\alpha\subseteq\cup_{1\le i\le n}A_i$ containing at most
one action per component, that is, $\alpha\cap\mc{A}_i$ is either empty or a 
singleton $\{a_i\}$. Thus, an interaction $\alpha$ can be put in the form
$\alpha=\{a_i\}_i{\in{I}}$ with $I\subseteq\{1,\cdots,n\}$ and $a_i\in\mc{A}_i$
for all $i\in{I}$. We denote by $\p{\alpha}$, the set of components \emph{participating}
in $\alpha$, that is, $\p{\alpha}=\{B_i\}_{i\in{I}}$.

\begin{definition}[Timed System]\label{def:comp}
  Given $n$ components $\tci{i}$ with $\Loc_i\cap\Loc_j=\emptyset, \ \A_i\cap
  \A_j=\emptyset,\text{ and }\X_i\cap\X_j=\emptyset$ for any $i\neq j$,
  the composition with respect to the interaction set $\gamma$, denoted by 
  $S=\gamma(B_1,\cdots,B_n)$, is defined by the timed component 
  $(\Loc,\loc_0,\X,\gamma,\E_{\gamma},\I)$ where:
  \begin{itemize}
    \item $\Loc=\Loc_1\times\cdots\times\Loc_n$
    \item $\loc_0=(\loc_{0_1},\cdots,\loc_{0_n})$,
    \item $\mc{X}=\mc{X}_1\cup\cdots\cup\mc{X}_n$,
    \item $\Inv{\loc}=\Invi{\loc_1}{1}\wedge\cdots\wedge\Invi{\loc_n}{n}$, for $\loc=
      (\loc_1,\cdots,\loc_n)$,
    \item $\mc{E}_{\gamma}$ is defined by:
  \end{itemize}
      \begin{myequation}
        \mc{E}_{\gamma}=\left\{
          \substack{\loc\transit{\alpha,g,r}\loc'\\
            \text{ for } \alpha=\{a_i\}_{i\in{I}}\in\gamma}\Big\lvert
        \substack{\loc=(\loc_1,\cdots,\loc_n)\in\Loc, \  
        \loc'=(\loc_1',\cdots,\loc_n')\in\Loc \\
        \text{ if } i\not\in{I},\ \loc_i'=\loc_i,\text{ and for }i\in{I},\
        \loc_i\transit{a_i,g_i,r_i}\loc_i'\text{ and } \\ 
        g=\bigwedge_{i\in{I}}g_i,\  r=\cup_{i\in{I}}r_i} \right\} 
      \end{myequation}
      
\end{definition}

In a composition $S$ of $n$ components $B_i$, an action $a_i$ can execute only as part
of an interaction $\alpha$ such that $a_i\in\alpha$, that is, along with the 
execution of all other actions $a_j\in\alpha$. This corresponds to the usual 
notion of multiparty interactions.
In practice, we do not explicitly build compositions of timed components as 
presented in Definition~\ref{def:comp}. We rather interpret their semantics by 
evaluating enabled interactions based on current states of components. 
\begin{property}[Semantics of a Composition]\label{pr:std_sem}
  Given a set of components $\{B_1,\cdots,B_n\}$ and an interaction set $\gamma$,
  the semantics of the composite component $S=\gamma(B_1,\cdots,B_n)$
  with respect to the set of interaction $\gamma$, is defined by 
  the timed transition system $\TTSg$ where:
  \begin{itemize}
    \item $\mc{Q}_g=\Loc\times\Val$ is the set of global states, where
      $\Loc=\Loc_1\times\cdots\times\Loc_n$ and $\mc{X}=\bigcup_{i=1}^n\mc{X_i}$.
      We write a state $q=(\loc,\val)$ where $\loc=(\loc_1,\cdots,\loc_n)\in\Loc$
      is a global location and $\val=(\val_1,\cdots,\val_n)\in\Val$ is a global 
      clock valuation. The initial state is $q_0=(\loc_0,0)$,
    \item $\gamma$ is the set of interactions,
    \item $\to_{\gamma}$ is the set of transitions defined by the rules:
    \begin{itemize}
      \item Interaction step: \\
      \end{itemize}
  \end{itemize}
  \vspace*{-1cm}
  \begin{align*}
    & \alpha=\{a_i\}_{i\in I}\in\gamma,\quad\forall i\in{I}.(\loc_i,\val_i)
    \transitb{a_i}{\gamma}(\loc'_i,\val'_i),\quad\forall i\notin{I}.
    (\loc_i,\val_i)=(\loc'_i,\val'_i)\\
    \cline{1-2}
   &\hspace{4.5cm}(\loc,\val)\transitb{\alpha}{\gamma}(\loc',\val')
 \end{align*}
  \vspace*{-1.5cm}
  \begin{itemize}
    \item[] 
      \begin{itemize}
      \item Time step:
    \end{itemize}
  \end{itemize}
  \vspace*{-5mm}
  \begin{align*}
    & d\in\realpos,\quad\forall i\in\{1,\cdots,n\}
    .\val_i+d\models\Invi{\loc_i}{i}\\
    \cline{1-2}
   &\hspace{2cm}(\loc,\val)\transitb{d}{\gamma}(\loc,\val+d)
 \end{align*}
\end{property}

To simplify notations, predicates defined on individual components $B_i$ are
straightforwardly interpreted on global states $(\loc,\val)$ of the composition
by considering the projection $(\loc_i,\val_i)$ of $(\loc,\val)$ on $B_i$.
For instance, $\al{\loc_i}$ evaluates to true on $(\loc,\val)$ iff
$\loc\in\Loc_1,\times\cdots\times\Loc_{i-1}\times\{\loc_i\}\times\Loc_{i+1}\times
\cdots\times\Loc_n$. Similarly, clock constraints of component $B_i$ are applied to clock 
valuation functions of the composition
by restricting $\val$ to clocks in $\mc{X}_i$ of $B_i$. This allows to write
the predicate $\enabled{\alpha}$, characterizing states $(\loc,\val)$ from which 
an interaction $\alpha=\{a_i\}_{i\in{I}}\in\gamma$ can be executed, as:
\begin{align}\label{eq:en}
  \enabled{\alpha}&=\bigwedge_{i\in{I}}\enabled{a_i},\\
                  &=\bigwedge_{i\in{I}}\bigvee_{\loc_i\in\Loc_i}\al{\loc_i}\wedge
                  \Phi(a_i,\loc_i),\label{eq:en_1}\\
                  &=\bigwedge_{\substack{\loc\in\Loc\\ 
                  \loc=(\loc_1,\cdots,\loc_n)}}\al{\loc}\wedge
                  \bigwedge_{\substack{i\in I\\ a_i\in\alpha }}\Phi(a_i,\loc_i)\label{eq:en_2}
\end{align}
Expression~\ref{eq:en_1} expresses the predicate $\enabled{\alpha}$ using location of individual
components whereas Expression~\ref{eq:en_2} formalizes it on global location configurations.
Notice that the above formulation of $\enabled{\alpha}$ corresponds to locations 
enumeration of all components participating in interaction $\alpha$. In practice,
we rather consider only a subset of locations $\Loc_{\alpha}\subseteq\Loc$, from
which the execution of $\alpha$ is possible. This corresponds to $\Pi_{i\in{I}}
|\Loc_{a_i}|$ possible configuration, where $\Loc_{a_i}\subseteq\Loc_i$ is a subset
of locations from which there exists a transition labeled by action $a_i\in\alpha$,
and $|\Loc_{a_i}|$ denotes the cardinality of $\Loc_{a_i}$, which is reasonably
small in practical examples but can be (at the worst case) equal to $\Pi_{i\in{I}}
|\Loc|$. The predicates $\enabledbackward{\alpha}$, 
$\enabledbackwardb{\alpha}{l}{u}$ and $\enabledforward{\alpha}$ becomes:

\begin{align*}
  &\enabledbackward{\alpha}\hspace{1.8mm}=\bigvee_{\substack{\loc\in\Loc\\
  \loc=(\loc_1,\cdots,\loc_n)}}\al{\loc}\wedge \backward\big(\bigwedge_{\substack{
    i\in{I}\\a_i\in\alpha}}\Phi(a_i,\loc_i)\big) \\
  &\enabledbackwardb{\alpha}{l}{u}=\bigvee_{\substack{\loc\in\Loc\\
  \loc=(\loc_1,\cdots,\loc_n)}}\al{\loc}\wedge \backwardp{l}{u}\big(\bigwedge_{
    \substack{i\in{I}\\a_i\in\alpha}}\Phi(a_i,\loc_i)\big) \\
  &\enabledforward{\alpha}\hspace{1.9mm}=\bigvee_{\substack{\loc\in\Loc\\
  \loc=(\loc_1,\cdots,\loc_n)}}\al{\loc}\wedge \forward\big(\bigwedge_{\substack{
    i\in{I}\\a_i\in\alpha}}\Phi(a_i,\loc_i)\big) \\
\end{align*}

Notice that for a clock constraint $c=c_1\wedge c_2$, we have:
\begin{displaymath}
  \diamond c = \diamond (c_1\wedge c_2) \neq \diamond c_1\wedge \diamond c_2
\end{displaymath}
with $\diamond\in\{\backward,\forward\}$.

The definitions of deadlocks and action-time-locks are also trivially extended
to composition of timed components. Deadlocks can be characterized as follows:
\begin{displaymath}
  \bigvee_{\loc=(\loc_1,\cdots,\loc_i)\in\Loc}\al{\loc}\wedge
  \big[\bigwedge_{\alpha\in\gamma}\neg\backward\big(\enabled{\alpha}\wedge
  \bigwedge_{1\le i\le n}\Invb{i}{\loc_i}\big)\big]
\end{displaymath}
and action-time-locks by:
\begin{displaymath}
  \big(\bigwedge_{\alpha\in\gamma}\neg\enabled{\alpha}\big)\wedge\big(
  \bigvee_{1\le i\le n}\bigvee_{\loc_i\in\Loc_i}\al{\loc_i}\wedge urg(\loc_i)\big)
\end{displaymath}

\subsection{Example}\label{exp:run}
\input{Figures/taskManager.tex}
Figure~\ref{fig:tm} depicts a timed system composed of four components $C$, $T_1$, $T_2$, 
and $R$.
Component $C$ represents a  controller that initializes then releases tasks $T_1$ and $T_2$.
Tasks use the shared resource $R$ during their executions.
To implement such behavior, we consider the following interactions between $C$, $R$, and 
$T_1$: $\alpha_1=\{init_0, init_1\}$,
$\alpha_3=\{start_0, start_1\}$, $\alpha_5=\{ take, process_1\}$, $\alpha_7 = 
\{free, end_1 \}$, 
and similar interactions $\alpha_2$, $\alpha_4$, $\alpha_6$, $\alpha_8$ for task $T_2$, 
as shown by connections on Figure~\ref{fig:tm}.
The controller is responsible for firing
the execution of each task. First, it non-deterministically initializes one
of the two tasks, i.e., executes $\alpha_1$ or $\alpha_2$, and then
releases it through interaction $\alpha_3$ or $\alpha_4$.
Tasks perform their processing independently of the controller, after being granted an access 
to the shared resource ($\alpha_5$ or $\alpha_6$).
When finished, a task releases the resource (interactions $\alpha_7$ or $\alpha_8$) and goes 
back to its initial location.
An example of execution sequence of this system is given below. Valuation $v$ of clocks $x$, 
$y$, and $z$ are represented as tuples 
$(\val(x),\val(y),\val(z))$:
\begin{displaymath}
\small{
\begin{split}
&((\loc_0^1,\loc_0^2,\loc_0^3,\loc_0^4),(0,0,0))\transit{26}_{\gamma}
((\loc_0^1,\loc_0^2,\loc_0^3,\loc_0^4),(26,26,26))\transit{\alpha_1}_{\gamma}
((\loc_1^1,\loc_1^2,\loc_0^3,\loc_0^4),(26,26,26))\transit{\alpha_3}_{\gamma}\\
&((\loc_0^1,\loc_2^2,\loc_0^3,\loc_0^4),(0,26,0))\transit{10}_{\gamma}
((\loc_0^1,\loc_2^2,\loc_0^3,\loc_0^4),(10,36,10))\transit{\alpha_5}_{\gamma}
((\loc_0^1,\loc_3^2,\loc_0^3,\loc_1^4),(0,36,10))\transit{2}_{\gamma}\\&
((\loc_0^1,\loc_3^2,\loc_0^3,\loc_1^4),(2,38,12))\transit{\alpha_2}_{\gamma}
((\loc_1^1,\loc_3^2,\loc_1^3,\loc_1^4),(2,38,12))
\end{split}
}
\end{displaymath}

\section{Timed Systems with Data}
Timed models introduced in the previous section focuses on the timing behavior of a given system.
In order to achieve a higher degree of expressiveness, we extend these models with data variables.
Data allows additional representations of complex behavior. Similarly to clock variables, they may
appear in the guards of transitions as additional conditions and may be updated when transitions
fire.

\begin{definition}[Guards on clocks and Data]\label{def:guard}
  Let $\mc{X}$ be a set of clock variables and $\mc{D}$ be a set of
  data variables. We denote by $\mc{G(X,D)}$ the set of guards induced
  by the following grammar:
  \begin{displaymath}
    g:=g_x \ | \ g_d \ | \ g_1\wedge g_2
  \end{displaymath}
  where $g_x\in\mc{C(X)}$, $g_d$ is a predicate on a subset of data variables
  of $\mc{D}$, and $g_1$, $g_2$ are guards over clocks and/or data variables.
\end{definition}

We extend the notion of valuation to data variables in the following manner:
\begin{itemize}
  \item Valuations assign values to data variable (in addition to clocks),
  \item The satisfaction of a valuation to a constraint is straightforwardly 
    extended to data variables,
  \item Data variables are insensitive to the progress of time, that is, for $k\in\mc{D}$ and
    $d\in\realpos$, $(v+d)(k)=(v)(k)$,
  \item Update operations are also defined for data variables using transfer functions

\end{itemize}

We use transfer functions to express update operations on data variables of $\D$. A transfer 
function $f:\V(\D)\to\V(\D)$ assigns to each variable $d\in\D$ a new value $f(v)$ based on the
current values of variables of $\D$. Additionally, given a set $\D'\supseteq\D$, applying $f$
on $\D'$ does not change the values of variables in $\D'\setminus\D$.

\begin{definition}[Timed Component with Data]\label{def:tce}
  A timed component with data is a tuple 
  $B_d=(\Loc,\loc_0,\X,\D,\A,\E,\{f_e\}_{e\in\E},\I)$ such that:
  \begin{itemize}
    \item $\Loc$, $\loc_0$, $\mc{X}$, $\mc{A}$ and $\mc{I}$ are defined as
      in Definition~\ref{def:tc},
    \item $\mc{D}$ is a finite set of data variables,
    \item $\mc{E}\subseteq\Loc\times(\mc{A}\times\mc{G(X,D)}\times2^{\X})\times\Loc$
    is a finite set of labeled transitions with an action, an extended guard, 
      and a set of clocks to be reset. For each transition $e\in\E$, we include the transfer
      function $f_e$ that updates elements of $\mc{D}$. 
  \end{itemize}
\end{definition}

\input{Figures/tcd.tex}
\begin{example}
  Component of Figure~\ref{fig:tcd} is an extended timed component with actions $a$, $b$ and $c$.
  The set of data is $\mc{D}=\{k\}$. It is used on the guards of the transitions labeled by
  $a$ and $c$ ($k\le 5$ and $k>5$ respectively). The update operations for both transitions
  are respectively $\{k:=k+10\}$ and $\{k:=k-3\wedge x:=0\}$.
\end{example}
\begin{remark}
  Notice that extending a timed component with data will result in a restriction
  of the behavior of the initial timed components since extending guards to data
  variables will only constrain the execution of transitions. Consequently,
  for a timed component with data $B_d=(\Loc,\loc_0,\X,\D,\A,\E,\{f_e\}_{e\in\E},\I)$, 
  the timed component $B=(\Loc,\loc_0,\X,\A,\E,\I)$ represents an abstraction of the latter. 
\end{remark}

\begin{definition}[Timed System with Data]\label{def:tce_sem}
  Given $n$ components $B_{d_i}=(\Loc_i,\loc_{0_i},\X_i,\D_i,\A_i,$\\$
  \E_i,\{f_e\}_{e\in\E_i},\I_i)$ and an interaction set $\gamma$,
  the composition $S_d=\gamma(B_{d_1},\cdots,B_{d_n})$ with respect to 
  the interaction set $\gamma$ is defined by the timed component 
  $(\Loc,\loc_0,\X,\D,\gamma,\E_{\gamma},\{f_e\}_{e\in\gamma},\I)$ where:
  \begin{itemize}
    \item $\Loc$, $\loc_0$, $\mc{X}$, $\gamma$ and $\mc{I}$ are defined as
      in Definition~\ref{def:comp}.
    \item $\mc{D}=\mc{D}_1\cup\cdots\cup\mc{D}_n$
    \item $\mc{E}_{\gamma}$ is straightforwardly extended by considering guards
      on data variables and the application of transfer functions on executions of interactions
  \end{itemize}
\end{definition}
The semantics of timed system with data extends the semantics of Property~\ref{pr:std_sem}
in the sense that an interaction takes place if the guards on data variables also evaluates
to $\true$. Also, interactions executions apply the underlying transfer functions of the
corresponding components transitions.

In what follows, when referring to timed component or timed system we imply by that 
extended timed component or extend timed systems, unless explicitly stated. 

\section{Verification of Timed Systems}\label{sec:2.4}
\subsection{Symbolic Reachability}

The semantics of timed systems as presented in Property~\ref{pr:std_sem} 
defines states as pairs of locations configurations and clock
valuations. By considering a continuous time domain ($\realpoz$ here), the 
resulting timed transition system yields an infinity of states.
Consequently, the usual practice is to rely on a symbolic representation
of states to make this state space finite.
A symbolic state is defined by a pair $(\loc,\xi)$ where $\loc$ is a location
and $\xi$ is a zone, a set of clock valuations defined by clock constraints
(as defined in~\ref{eq:cc}). Consequently, the set of reachable states of a 
timed component B (system) can be put on the form:
\begin{displaymath}
  Reach(B)=\bigvee_{j\in{J}}\al{\loc_j}\wedge\xi_j
\end{displaymath}

Given a timed component with data $B_d$ and its abstraction of data $B$, we have
$Reach(B_d)\subseteq Reach(B)$. This means that $Reach(B)$ can be used as an over-approximation
of the reachable state of the timed component with data $B_d$.
\subsection{Compositional Verification}

Standard verification techniques such as model checking are based on explicit 
exploration and exhaustive enumeration of all the reachable symbolic states 
of a given system. The main issue with this 
method is the combinatorial explosion when considering large scale systems.
Compositional verification have been introduced to cope with state explosion
problem, and thus achieves scalability when verifying large scale systems.
This approach is based on the concept of divide-and-conquer in order to 
break up the verification
problems into smaller subsequent problems. Compositional verification have been 
extensively studied under different manners e.g., assume-guarantee 
reasoning~\cite{Lamport77,Owl76}, 
contract-based verification~\cite{contract1,contract2},  
deductive verification
~\cite{deductive}, etc.  
In this thesis, we choose to use a deductive compositional verification 
method that exploits compositionality for analysis of timed systems 
using \emph{invariants}. Invariants are symbolic approximations of the set 
of reachable states of the system as opposed to the exact 
reachability analysis in model-checking. The key principle of this 
approach is the computation of a global invariant as the conjunction of
other invariants (components invariants, interaction invariants, etc.).



%:\emph{(i)} local invariants of 
%the underlying components, \emph{(ii)} an interaction 
%invariant deduced from the synchronizations 
%(interactions) between components, and \emph{(iii)} an additional invariant 
%capturing the 
%global timing aspects and relating clocks of different 
%components.

Let $S=\gamma(B_1,\cdots,B_n)$ be a system composed of $n$ timed components $B_i$ 
synchronizing through the interaction 
set $\gamma$, and let $\psi$ be a property of interest. Assuming that 
$GI(S)$ is the global invariant for this system, 
the verification rule of $\psi$ can be intuitively
written as follows:

\begin{align*}
  &\hspace{1cm}GI(S)\Rightarrow\psi\\
    \cline{1-2}
   &\gamma(B_1,\cdots,B_n)\models\square\psi
\end{align*}
where the notation \enquote{$\gamma(B_1,\cdots,B_n)\models\square\psi$} is to 
be read as \enquote{$\psi$ holds in every reachable state of the
composition $\gamma(B_1,\cdots,B_n)$}.

Usually when verifying timed systems, the common practice is to verify the system 
against some \emph{given property} (safety property, liveness, deadlock, etc.).
These properties allows to assert that a given model satisfies the specifications.
We use satisfiability checking to verify that the global 
invariant of a system 
implies properties of interest. Particularly, for a system S characterized 
by the global invariant $GI(S)$, and 
given a property $\psi$, we check the unsatisfiability of 
$GI(S)\wedge\neg\psi$. 



