\begin{acknowledgements}
\begin{otherlanguage}{french}
\ackfont
\large
\lettrine[lines=3,loversize=0.2]{\color{BrickRed}F}{} irstly, I would like to express my 
sincere gratitude to my advisor Prof. Saddek Bensalem for giving me the opportunity to carry 
out my PhD at the Verimag laboratory, for providing ideal research
conditions and for all his trust and support during this period. 
I would like to deepely thank my co-advisor Dr. Jacques Combaz for his continuous support, 
patience, motivation, and immense knowledge. His guidance as well as the tremendous 
amount of time spent with me all along my PhD are pricesless.
I want to thank Dr. Marius Bozga for his precious suggestions, advices and for 
sharing with me all his knowledge in formal methods.

I would like to thank Prof. Stavros Tripakis and Prof. Oleg Sokolsky for accepting to reviw this
thesis. Many thanks to Prof. Florence Maraninchi, Prof. Patricia Bouyer-Decitre for accepting 
to be part of the jury.
  
This journey would not have been possible without my family. To my parents, who always favored
their children without asking or expecting anything in return. Thank you for all your 
encouragement, assistance and for the emotional and financial supports. Special thanks to my 
sisters for all the time and efforts spent while trying to cheer me up during difficult moments.
Special thoughts to my maternal grandmother who passed away during my first year of PhD.
  
My sincere thanks to Lotfi Mediouni for all the time spent together during coffee breaks, for
his help and especially for his optimism (I think I have never seen a person that optimistic
in my life). Special thanks to all my colleagues Hosein Nazarpour, Ayoub Nouri, Rany Kahil, 
Rim El-Belouli and all my Verimag collegues.

Last but not least, thanks you Anne for her continuous support, and to her familly
for all their kindness and the endless invitations to family dinners, christmass and every other
occasion. They made me feel I am home.
     
\end{otherlanguage}
\end{acknowledgements}

